---
title: "GEOG 176A: Intro to GIS"
author: "[Ian Hinds](https://hindstein.github.io/GEOG176A/)"
subtitle: 'Lab 03: Distances and the Border Zone'
output:
  html_document:
    theme: journal
---

## Libraries
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# SPDS
library(tidyverse)
library(sf)
library(units)

# Data
library(USAboundaries)
library(rnaturalearthdata)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

## Background
In this lab, 4 main skills are covered:

* 1. __Ingesting/Building__ _sf_ objects from R packages and CSVs. (Q1)
* 2. __Manipulating__ geometries and coordinate systems (Q2)
* 3. __Calculaing__ distances (Q2)
* 4. __Building__ maps using ggplot (Q3)

## Question 1:
For this lab, 3 datasets are needed.

* 1. Spatial boundaries of continental USA states (1.1)
* 2. Boundaries of Canada, Mexico, and the United States (1.2)
* 3. All USA cities (1.3)

### _1.1 Define a Projection_
For this lab we want to calculate distances between features, therefore we need a projection that perserves distance at the scale of CONUS. For this, we will use the North American Equidistant Conic:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
eqdc = '+proj=eqdc +lat_0=40 +lon_0=96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs' 
```
Unfortunately this projection does not have a EPSG code to reference so we must ue the PROJ4 string above. Take a moment and describe this projection based on the parameters:




### _1.2-Get USA state boundaries_
In R, USA boundaries are stored in the  'USAboundaries' package.

