---
title: "GEOG 176A: Intro to GIS"
author: "[Ian Hinds](https://hindstein.github.io/GEOG176A/)"
subtitle: 'Lab 05: Rasters & Remote Sensing'
date: "4 September 2020"
output:
  html_document:
    theme: journal
---

******

## Background
On September 26, 2016 at 11:47 a.m. U.S. Central Daylight Time (16:47 UTC) the Cedar and Wapsipinicon rivers in Iowa surged producing a flood wave that breached the river banks. The water level of the Cedar River measured ~20 feet — 8 feet above flood stage—near the city of Cedar Rapids.

The water level continued to rise until it peaked at ~22 feet on September 27. This event had only been exceeded once, in June 2008, when thousands of people were encouraged to evacuate from Cedar Rapids, the second-most-populous city in Iowa.

In this lab we are interested in the impacts in Palo Iowa because it is up stream of Cedar Rapids, contains a large amount of farm land, and does not have a forecast location to provide warning.

We will use the raster package and our understanding of raster data and categorization to create flood images using mutliband Landsat Imagery, thresholding, and classification methods.

### Libraries
```{r message=FALSE, warning=FALSE}
library(raster) # Raster Data handling
library(tidyverse) # Data Manipulation
library(getlandsat) # keyless Landsat data (2013-2017)
library(sf) # Vector data processing
library(mapview) # Rapid Interactive visualization
```

Remote Sensing/Image Analysis begins with usually the same steps:

- 1. Identifying an area of interest (AOI)
- 2. Identifying and downloading the relevant images or products
- 3. Analyzing the raster products

*****

## *Question 1 (AOI Definition)*

Using uscities.csv, filter to Palo, Iowa and create a 5 kilometer buffer.

- Read in the csv data
- Structure it as a spatial (sf) feature by defining the coordinate fields and CRS
- Filter to only include Palo, Iowa
- Transform the projection into a CRS appropriate for meter-based measurements

From the Palo, Iowa feature:

- generate a 5 kilometer (5,000 meter) buffer around the point using st_buffer
- find the bounding box of the buffered region using st_bbox
- Convert that bbox object into as sfc then sf object

This region defines the AOI for this analysis.
```{r}






```

