---
title: "GEOG 176A: Intro to GIS"
author: "[Ian Hinds](https://hindstein.github.io/GEOG176A/)"
subtitle: 'Lab 02: COVID-19 Pandemic'
output:
  html_document:
    theme: journal
---

## Libraries
* *tidyverse*
* *knitr*
* *readxl*
* *zoo*

```{r include=FALSE}
library(tidyverse)
library(knitr)
library(readxl)
library(zoo)
```

## COVID Data URL (NYTimes)
https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv

```{r include=FALSE}
library(tidyverse)
url = 'https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv'
covid = read_csv(url)
```
## Population Data (USDA)
https://www.ers.usda.gov/webdocs/DataFiles/48747/PopulationEstimates.xls?v=2561.3

## Question 1
As it stands, the California Department of Public Health maintains a watch list of counties that are being monitored for worsening coronavirus trends. There are six criteria used to place counties on the watch list:

* Doing fewer than 150 tests per 100,000 residents daily (over a 7-day average)

* More than 100 new cases per 100,000 residents over the past 14 days…

* 25 new cases per 100,000 residents and an 8% test positivity rate

* 10% or greater increase in COVID-19 hospitalized patients over the past 3 days

* Fewer than 20% of ICU beds available

* Fewer than 25% ventilators available

**Of these 6 conditions, we will monitor condition number 2.** 

To do this job well, you should set up a reproducible framework to communicate the following in a way that can be updated every time new data is released:

* cumulative cases in the 5 worst counties
* total NEW cases in the 5 worst counties
* A list of safe counties
* A text report describing the total new cases, total cumulative cases and number of safe counties.

__Cumulative cases in the 5 worst counties to date:__
```{r echo=FALSE, message=FALSE, warning=FALSE}
worst_counties = covid %>%
  filter(state == "California", date == max(date)) %>%
  group_by(county) %>%
  summarize(cases = max(cases, na.rm = TRUE)) %>%
  ungroup() %>%
  slice_max(cases, n=5)
knitr::kable(worst_counties, 
             caption = "Counties with the Most Cases",
             col.names = c("County", "Cases"),
             color = "black")
```

__Total NEW cases in the 5 Worst Counties Today__
```{r message=FALSE, warning=FALSE, include=FALSE}
worst_counties = covid %>%
  filter(state == "California") %>%
  group_by(county) %>%
  mutate(newCases = cases - lag(cases)) %>%
  ungroup() %>%
  filter(date == max(date))

(most_new_cases = worst_counties %>%
      slice_max(newCases, n = 5) %>%
      select(county, newCases))
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(most_new_cases, 
             caption = "New Cases", 
             col.names = c("County", "Cases"))
```


```{r include=FALSE}
#Open population estimate excel
library(readxl)
popest <- read_excel("C:/Users/Ianhi/OneDrive/Desktop/github2/labs/data/popest.xls")
View(popest)
```
```{r eval=FALSE, include=FALSE}
# Join COVID and Population
#full_join(covid, popest, by = "California") %>%
#group_by(county, date) %>%
#summarize()
  
```


## Question 2
In this question, we are going to look at the story of 4 states and the impact scale can have on data interprtation. The states include: New York, California, Louisiana, and Florida.

We will make a faceted bar plot showing the number of daily, new cases at the state level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Group/Summarize county level to state level, filter it to the four states of interest
state.of.interest = "New York"

covid %>%
  filter(state == state.of.interest) %>%
  group_by(date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),
         roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = newCases), col = NA, fill = "gold") +
  geom_line(aes(y = roll7), col = "darkblue", size = 1) +
  ggthemes::theme_wsj() +
  labs(title = paste("New Reported cases by day in", state.of.interest)) +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 14, face = 'bold')) +
  theme(aspect.ratio = .5)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
state.of.interest = "California"

covid %>%
  filter(state == state.of.interest) %>%
  group_by(date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),
         roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = newCases), col = NA, fill = "gold") +
  geom_line(aes(y = roll7), col = "blue", size = 1) +
  ggthemes::theme_wsj() +
  labs(title = paste("New Reported cases by day in", state.of.interest)) +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 14, face = 'bold')) +
  theme(aspect.ratio = .5)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
state.of.interest = "Louisiana"

covid %>%
  filter(state == state.of.interest) %>%
  group_by(date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),
         roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = newCases), col = NA, fill = "gold") +
  geom_line(aes(y = roll7), col = "darkblue", size = 1) +
  ggthemes::theme_wsj() +
  labs(title = paste("New Reported cases by day in", state.of.interest)) +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 14, face = 'bold')) +
  theme(aspect.ratio = .5)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
state.of.interest = "Florida"

covid %>%
  filter(state == state.of.interest) %>%
  group_by(date) %>%
  summarise(cases = sum(cases)) %>%
  mutate(newCases = cases - lag(cases),
         roll7 = rollmean(newCases, 7, fill = NA, align="right")) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = newCases), col = NA, fill = "gold") +
  geom_line(aes(y = roll7), col = "darkblue", size = 1) +
  ggthemes::theme_wsj() +
  labs(title = paste("New Reported cases by day in", state.of.interest)) +
  theme(plot.background = element_rect(fill = "white"),
        panel.background = element_rect(fill = "white"),
        plot.title = element_text(size = 14, face = 'bold')) +
  theme(aspect.ratio = .5)
```

>Conclusion: States with higher populations represented larger spikes in daily cases. These states are Florida, California, and New York. Louisiana, containing a much smaller populous, experiences a less extreme variance in case frequency. Scaling by population would make the larger populations look more similar, and the smaller populations more different.

## Question 3:

>Metropolitan environments and other largely populated areas throughout the U.S. have seen drastic spikes in case numbers in a short time span. The inconsistent practice of wearing a mask and social distancing has negativvely affected some communites, especially those who are at risk of being a hot spot. COVID-19 can spread easiest when physical and close contact is exchanged with individuals, therefore larger communities have seen the greatest number of cases.


